Make any stupid plot you want

### 2021-04-29 draft/sections/intro.tex
reworked viz section

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Building block visualization libraries implement the functions that encode data as visual representations. For example, functions that map latitude and longitude into positions on a graphical representation of a map and functions that map precipitation and station name to colors. Visualization libraries also provide functions that composite these encoding functions into attributes of a visual element, such as a point, and we call these functions \textit{artists}.  The encoders and \textit{artists} are reusable, so position encoders can also be used to to translate date and time to positions, which can then be composited into a line by a \texttt{line} function. 
    }
    \label{fig:intro:artist}
\end{figure}

Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. We define visualization components to be structure preserving maps from data to visual representations. As illustrated in \autoref{fig:intro:artist}, visualizations map data into graphical elements such as points and lines and map individual data components (columns) into components of the visualization such as position or color. For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors. A \texttt{point} function composites these encodings into attributes of a point on screen. The \texttt{line} function composites position and color encodings into a line, specifically encoders that convert date and precipitation to x and y positions and a map from station name to color. We identify the structure these maps must preserved as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.

We introduce a model of visualization components based on these \textit{continuity} and \textit{equivariance} constraints and use this model to develop a design specification. Our specification is targeted at building block libraries, as defined by Wongsuphasawat\cite{wongsuphasawatNavigatingWideWorld2021}, because they are the visualization tools that provide independent functions that map components of data to visual representations.  Just as physical building blocks can be arranged in an almost infinite number of ways, solely constrained by the size and shape of the blocks, we propose that the only restriction on the composition of building block library components are \textit{continuity} and \textit{equivariance}. We adopt a functional model for our proposed architecture because independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}. Doing so means the functional architecture can be evaluated for correctness, the resulting code is likely to be shorter and clearer, and the architecture is well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.

This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library. We even name these compositing functions \texttt{point} and \texttt{line} in \autoref{fig:intro:artist} \textit{artists} as a nod to the somewhat analogous object in the Matplotlib architecture.  One of the most widely used visualization libraries in Python, since 2002 new components and features have been added in a some what adhoc, sometimes hard to maintain, manner. Particularly, each new component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is then expressed in the API for that component. In turn, this yields an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}. This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g. shape or color, are supported. We propose that these issues can be ameliorated via a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints. We evaluate our functional model by implementing new artists in Matplotlib that are specified via \textit{equivariance} and \textit{continuity} constraints. We then use the common data model introduced by the model to demonstrate how plotting functions can be consolidated in a way that makes clear whether the difference is in expected data structure, visual component encoding, or the resulting graphic.

\section{Background}
There are many formalisms of the notion that visualization is structure preserving maps from data to visual representation, and many visualization libraries that attempt to preserve structure in some manner; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level to propose a new model of the constraints visual transformations must satisfy such that they can be composed to produce visualize representations that can be considered equivalent to the data being represented.

\subsection{Structure:}
\label{sec:intro:structure}
 The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.

\begin{figure}[H]
        \includegraphics[width=1\columnwidth]{figures/intro/equivariant.png}
        \caption{The data in blue is scaled by a factor of two, yielding the data in orange. To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
        \label{fig:intro:equivariance}
\end{figure}
Bertin proposes that there are classes of visual encodings-such as position, shape, color, and texture-that when mapped to from specific types of measurement, quantitative or qualitative, will preserve the properties of that measurement type. For example, in \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from blue to orange data and lines. The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}. The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}. Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups. As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitely include continuity.

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/continuity.png}
    \caption{The line plot does not preserve \textit{continuity} because it implies that the average temperature at each station lie along a 1D continuous line, while the bar plot preserves \textit{continuity} by representing the average temperatures at each station as the discrete values they are.}
    \label{fig:intro:continuity}
\end{figure}
Bertin proposes that the visual encodings be composited into graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark. In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval. But, when the continuity is preserved, as in the bar chart, then the graphic has not introduced new structure into the data.

\begin{mdframed}[roundcorner=10pt, frametitle=Structure, frametitlerule=true, frametitlebackgroundcolor=gray!10]
    \begin{description}
        \item[\textbf{continuity}] How records in the dataset are connected to each other, e.g. discrete rows, neworked nodes, points on a continuous surface
        \item[\textbf{equivariance}] if an action is applied to the data or the graphic--e.g. a rotation, permutation, translation, or rescaling-- there must be an equivalent action applied on the other side of the transformation. 
    \end{description}
\end{mdframed}

The notion that a graphic should be equivalent to the data has been expressed in a variety of ways. Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis} states that a visualization is easier to understand when the properties of the visualization match the properties of the data. This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}. expressing, as defined by Mackinlay, is a measure how much of the mathematical structure in the data that can be expressed in the visualizations; for example that ordered variables can be mapped into ordered visual elements. We propose that a graphic is an equivalent representation of the data when \textit{continuity} and \textit{equivariance} are preserved.

>>>>

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Building block visualization libraries implement independent reusable functions that map data to visual representations. For example, position encoding functions map the latitude and longitude values in the table (a) to locations on the map (b). These same encoding functions map dates and precipitation (a) to x and y positions in the timeseries plot (c). Color encoding functions map the precipitation (a) to the color of the point in the map (a) and color encoding functions map names of weather stations (a) to colors in both the map (a) and timeseries (b). Functions, which we call \textit{artists}, compose these encoding functions into attributes of visual elements. The \texttt{point} artist transforms the output of the position, facecolor, and edgecolor maps into attributes of each point in the map (a), while the \texttt{line} artist transforms the output of the position and color maps into attributes of each line in the timeseries plot (c).}
    \label{fig:intro:artist}
\end{figure}

Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. We define visualization components to be structure preserving maps from data to visual representations. As illustrated in \autoref{fig:intro:artist}, visualizations map weather station precipitation data (a) into graphical elements such as points (b) and lines (c) and map individual data components (columns) into components of the visualization such as position or color. For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors. A \texttt{point} function composites these encodings into attributes of a point in the map (a). The \texttt{line} function composites encoders that convert station name to color and date and precipitation to x and y positions into a piece of a line in the timeseries plot (c). We identify the structure these maps must preserve as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.

We introduce a model of visualization components based on these \textit{continuity} and \textit{equivariance} constraints and use this model to develop a design specification. Our specification is targeted at building block libraries, as defined by Wongsuphasawat\cite{wongsuphasawatNavigatingWideWorld2021}, because they are the visualization tools that provide independent functions that map components of data to visual representations.  Just as the number of ways to arrange physical building blocks is solely constrained by the size and shape of the blocks, we propose that the only restriction on how building block library components can be composed are that the compositions preserve \textit{continuity} and \textit{equivariance}. Independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}, so we propose a functional architecture for our model. Doing so means the functional architecture can be evaluated for correctness, the resulting code is likely to be shorter and clearer, and the architecture is well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.

This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library. One of the most widely used visualization libraries in Python, new components and features have been added in an organic, sometimes hard to maintain, manner since 2002.  In Matplotlib, every component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is expressed in the API for that component. This leads to an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}. This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g. shape or color, are supported. We propose a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints. We evaluate our functional model by implementing new artists in Matplotlib that are specified via \textit{equivariance} and \textit{continuity} constraints. We then use the common data model introduced by the model to demonstrate how plotting functions can be consolidated in a way that makes clear whether the difference is in expected data structure, visual component encoding, or the resulting graphic.

\section{Background}
There are many formal models of visualization as structure preserving maps from data to visual representation, and many implementations of visualization libraries that preserve this structure in some manner. This work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level. We propose a data structure agnostic model of the constraints visual transformations must satisfy such that they can be composed to produce \textit{equivariant} and \textit{continuity preserving} visual representations.

\subsection{Structure:}
\label{sec:intro:structure}

\begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
        \caption{The precipitation in blue is scaled by a factor of 10, yielding the precipitation in orange in the table (a). This scaling factor is represented by the green arrows in (a). The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the green fill. Because the scalings are equivariant, the orange precipitation timeseries is 10x the blue precipitation timeseries. The red line is an example of \textit{equivariance} not being preserved, as it is not 10x the the y value of the blue precipitation timeseries.}
      \label{fig:intro:equivariance}
\end{figure}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}. Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data. For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in the table (a) is scaled by a factor of 10. This scaling action is represented by the green arrows that translate the precipitation into the scaled values in orange. For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b). The precipitation in orange is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the orange line is 10x the y-values of the blue precipitation timeseries. The green fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a). In contrast, the red line is not an \textit{equivariant} representation to the 10x scaling because the red y-values are not 10x the blue y values. The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}. The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}. Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups. As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitely include continuity.

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/continuity.png}
    \caption{The bar chart (b) and line plot (c) are different visual encodings of the precipitation data in the table (a). The bar chart preserves the \textit{continuity} of the stations by encoding the discrete station data as discrete bars (b). In contrast, the line plot (c) does not preserve this discrete \textit{continuity} because it connects the total temperature at each station by drawing a line through each point. Doing so implies that the total temperatures are points along a 1D continuous line, whereas there is no connectivity between the stations or their corresponding rows in the table.}
    \label{fig:intro:continuity}
\end{figure}
The notion that data is equivalent to visual representations when structure is preserved serves as the basis for visualization best practices. When \textit{continuity} is preserved, as in the bar chart (b) in \autoref{fig:intro:continuity}, then the graphic has not introduced new structure into the data.  In \autoref{fig:intro:continuity}, the line plot (c) does not preserve \textit{continuity} because the line connecting the total precipitation of the stations implies that the stations are connected to each other along a 1D continuous line. Bertin asserted that \textit{continuity} was preserved by choosing graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark. Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis} states that a visualization is easier to understand when the properties of the visualization match the properties of the data. This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}. Expressivity, as defined by Mackinlay, is a measure how much of the mathematical structure in the data can be expressed in the visualizations; for example that ordered variables can be mapped into ordered visual elements. We generalize these different codifications of structure preserving encodings, proposing that a graphic is an equivalent representation of the data when \textit{continuity} and \textit{equivariance} are preserved.

\begin{mdframed}[roundcorner=10pt, frametitle=Structure, frametitlerule=true, frametitlebackgroundcolor=gray!10]
    \begin{description}
        \item[\textbf{continuity}] How elements in the dataset are connected to each other, e.g. discrete points, networked nodes, points on a continuous surface
        \item[\textbf{equivariance}] if an action is applied to the data or the graphic--e.g. a rotation, permutation, translation, or rescaling-- there must be an equivalent action applied on the other side of the transformation. 
    \end{description}
\end{mdframed}
